---
title: "Heli-attack times"
author: "Michael Storey, MichaelJones, Mitchell Stephen, Michael May"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
---

```         
To Do:
  [x] Demonstration-quality waterpoints layer
  [ ] High-quality waterpoints/reservoir/lakes layer
  [ ] Classify water points (long line, medium heavy etc.)
  [ ] Improve helicopter travel model
  [x] Exclude demand points in the ocean/NSW
  [x] Improve leaflet map usability by converting point data to raster
  [ ] Erase raster values outside Gippsland footprint
  [x] Allow user to query map data
  [ ] Produce summary statistics for region, districts and workcentres
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#set local library path so R knows where to look. You will also need to run this line below before downloading new packages, just due to some cenitex issues.
#.libPaths("D:/R/R-Packages")

# load necessary libraries
  library(sf)
  library(dplyr)
  library(leaflet)
  library(leaflet.extras)
  library(htmltools)
  library(spatstat)
  library(leafem)
  library(raster)

```

### Setup

-   Define all inputs and variables

Remainder of script should run without requiring modification

```{r Chunk 1: Setup}
knitr::opts_chunk$set(cache = TRUE)

# path locations
  output_path <- "C:/Projects/20230829_HeliWaterPoints/waterpoints/outputs"
  input_path <- "C:/Projects/20230829_HeliWaterPoints/waterpoints/data/"

# shapefile names
  dem_file        <- "DEM_SE_Tas_mosaic.tif"
  wp_combined     <- "Waterpoints_Combined_20230912.shp"
  wp_new          <- "potential_dam_locations_14052020.shp"
  demand_points   <- "demandpoints_1kmIgnitionGrid.shp"
  districts       <- "LF_DISTRICT.shp"

# construct full file paths
  dem <- file.path(input_path, dem_file)
  #wp_private_land_full_path <- file.path(input_path, wp_private_land)
  #wp_public_land_full_path <- file.path(input_path, wp_public_land)
  wp_combined_full_path <- file.path(input_path, wp_combined)
  wp_new_full_path <- file.path(input_path, wp_new)
  demand_points_full_path <- file.path(input_path, demand_points)
  districts_full_path <-file.path(input_path, districts)

# helicopter speeds in kilometers per hour
  all_empty_kmh <- 148
  full_ascend_kmh <- 74
  full_level_kmh <- 148 #same as full descending

# path sample density (metres between samples from reservoirs to points)
  line_sample_point_dist <- 200

# choose whether to include new dams (1) or not (0). Don't use 1 under a 1k grid, will take too long.
  include_new_wp <- 0

# filter for one or more districts
  selected_districts <- ('DISTRICT_N == \"TAMBO\" | DISTRICT_N == \"MACALISTER\" | DISTRICT_N ==\"SNOWY\" | DISTRICT_N == \"LATROBE\"')

```

### Create demand grid

-   Load shapefiles
-   Transform coordinate reference system (CRS),and define the variable.
-   Create a point grid across region (ie encompassing the reservoirs of interest).
-   Plot grid points black and water in blue

```{r Chunk 2: Create demand grid, echo=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(cache = TRUE)

# load and project district shapefile
  district_poly <- st_read(districts_full_path, quiet = TRUE) %>%
    st_transform(3111, quiet = TRUE) %>%
    filter(DISTRICT_N == "TAMBO" | DISTRICT_N == "MACALISTER" | DISTRICT_N =="SNOWY" | DISTRICT_N == "LATROBE")

# load and project water points shapefile
  wp_combined_poly <- st_read(wp_combined_full_path, quiet=TRUE) %>% st_transform(3111, quiet=TRUE)
  
# load and project candidate sites for new helidams
  new_dams<-st_read(wp_new_full_path, quiet=TRUE) %>% st_transform(3111, quiet=TRUE) 
  #st_crs(new_dams)<-3111
  #point_grid<- water_poly %>% st_make_grid(point_grid_size,what='centers') %>% st_as_sf()

# Read and transform the demand points to a common CRS (e.g., 3111)
  demand_points <- st_read(demand_points_full_path, quiet = TRUE) %>% st_transform(3111, quiet=TRUE) 

# Use st_intersection to find points within polygons
  point_grid <- st_intersection(demand_points, district_poly)
  
  water_points<-st_cast(wp_combined_poly,"POINT")

# add new dams to water points if include_new_dams variable set to 1
  if(include_new_wp == 1){water_points<-st_union(water_points, new_dams) %>% st_cast("POINT")}

```

### Create traversal lines

-   Draw a line from every point in grid to its closest water source.

```{r Chunk 3: Create traversal lines, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(cache = TRUE)

# create lines from every grid point to nearest water point
water_points$wp_id <- seq.int(nrow(water_points))
water_points$geom_start <- water_points$geometry
point_grid$dp_id <- seq.int(nrow(point_grid))
point_grid <- st_join(point_grid, water_points, join=st_nearest_feature)
near_lines <- nngeo::st_connect(point_grid$geometry, point_grid$geom_start, k=1, progress = FALSE)
point_grid$line <- st_as_sf(near_lines) 


```

### Embed DEM into traversal lines

-   For every line from point grid to water, densify line (make vertex every 100 metres) and turn vertices into points (transferring line id to new point feature class)
-   Sample the DEM for every line vertex point, and select the point with the maximum elevation.
-   Sample the DEM for elevation of start point (where fire would be) and end point (water point)
-   Calculate the difference in elevation between start, end, and max elevation

```{r Chunk 4: Embed DEM, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(cache = TRUE)

# densify lines (make vertices every <line_sample_point_dist> metres)
  lines_dense <- point_grid$line %>% smoothr::densify(max_distance=line_sample_point_dist)
  lines_dense$dp_id <- point_grid$dp_id
  lines_vertices <- st_cast(lines_dense, "POINT", do_split = TRUE, warn = FALSE)

# sample dem for peak elevation
  raster_dem<-raster::raster(dem)
  
  lines_vertices$dem_path <- raster::extract(raster_dem, st_transform(lines_vertices, 4326, quiet=FALSE))
  max_dem_vertex <- lines_vertices %>%
    group_by(dp_id) %>%
    arrange(desc(dem_path)) %>%   # Arrange by dem_path in descending order
    slice(1) %>%                  # Take the first row within each group
    ungroup()
  point_grid$elev_peak <- max_dem_vertex$dem_path
  point_grid$geom_peak <- near_lines$start_geom <- st_transform(max_dem_vertex$x, 3111)

# sample dem for start and end points
  point_start <- st_as_sf(point_grid$geom_start)
  point_end <- st_as_sf(point_grid$geometry)
  point_peak <- st_as_sf(point_grid$geom_peak)
  
  point_grid$elev_start <- raster::extract(raster_dem,st_transform(point_start, 4326))
  point_grid$elev_end <- raster::extract(raster_dem,st_transform(point_end, 4326))

# calculate height differences
  point_grid$start_peak_diff <- as.double(point_grid$elev_peak) - as.double(point_grid$elev_start)
  point_grid$end_peak_diff <- as.double(point_grid$elev_peak) - as.double(point_grid$elev_end)


```

### Draw test plots

-   Plot 1 illustrates the demand grid and water point locations. Grid points are black circles and water is blue
-   Plot 2 adds traversal lines from demand grid points to nearest water point (brown lines)
-   Plot 3 includes the highest point vertex in red

```{r Chunk 5: Draw test plots, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(cache = TRUE)

# draw test plot of water point locations
  plot(st_geometry(point_grid), cex=0.1)
  plot(st_geometry(point_grid$geom_start), cex=0.1, col='blue', add=TRUE)

# draw test plot with lines between demand points and nearest water point
  plot(st_geometry(point_grid))
  plot(st_geometry(point_grid$geom_start), cex=0.1,col='blue', add=TRUE)
  plot(st_geometry(point_grid$line), cex=0.5, col="yellow", add=TRUE)

# draw test plot including max DEM point in purple
  plot(st_geometry(wp_combined_poly))
  plot(st_geometry(point_grid), cex=0.6, col='red', add=TRUE)
  plot(st_geometry(point_grid$geom_start), cex=0.2, col='blue', add=TRUE)
  plot(st_geometry(point_grid$line), cex=0.2, col="green", add=TRUE)
  plot(st_geometry(point_grid$geom_peak), cex=0.4, col = 'purple', add=TRUE)

```

### Calculate helicopter travel times

-   Calculate distance (2D) between:
    -   water point and demand point
    -   water point and highest elevation point
    -   highest elevation point and demand point
-   Adjust distances to account for changes in elevation
-   Calculate travel times assuming:
    -   74 km/h speed when ascending full
    -   148 km/h for other states (empty, level flight, descending)
    -   2 minutes for fill time
-   Summarise travel times using two different methods:
    -   Helicopter ascends vertically, flies level, then descends vertically
    -   Helicopter ascends at an angle between the water point and highest point; descends at an angle between the highest point and demand point
-   Create rasters using IDW (distance-weighted smoothing) for efficient maps

```{r Chunk 6: Helicopter travel times, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(cache = TRUE)

### NOTE:
### I've turned this into an inefficient for-loop because I had to work around a very large (73+GB) vector produced
### by the original code. And while creating the work-around, I came across the zero-value errors, which required debugging.
###
### This whole section could be improved and sped up by shifting to vector logic.
### But I want to retain it for now so we can implement a semi-realistic helicopter model (this one is garbage).

# Initialize vectors to store the calculated distances and inclines
  distance_t <- numeric(length(nrow(point_grid)))
  distance_s <- numeric(length(nrow(point_grid)))
  distance_e <- numeric(length(nrow(point_grid)))
  distance_s_3d <- numeric(length(nrow(point_grid)))
  distance_e_3d <- numeric(length(nrow(point_grid)))
  incline_s <- numeric(length(nrow(point_grid)))
  incline_e <- numeric(length(nrow(point_grid)))

# Loop through each row and calculate distances and inclines
  numrows <- nrow(point_grid)
  pb <- progress_estimated(numrows)
  cat("Analysing", numrows, "transects\n")
  for (i in 1:numrows) {
    pb$tick()$print()
    # Calculate distances
    distance_t[i] <- as.numeric(st_distance(point_start[i, ], point_end[i, ]))
    distance_s[i] <- as.numeric(st_distance(point_start[i, ], point_peak[i, ]))
    distance_e[i] <- as.numeric(st_distance(point_peak[i, ], point_end[i, ]))
    
    # Debugging
    #cat("point:", i, "of", numrows, "\n")
    # cat("distance_t:", distance_t[i], "\n")
    # cat("distance_s:", distance_s[i], "\n")
    # cat("distance_e:", distance_e[i], "\n")
    # cat("start_peak_diff:", point_grid$start_peak_diff[i], "\n")
    # cat("end_peak_diff:", point_grid$end_peak_diff[i], "\n")
    
    # Check for zero distances and handle them by assigning zero to distance_s_3d and incline_s
    if (distance_s[i] == 0) {
      distance_s_3d[i] <- 0
      incline_s[i] <- 0
    } else if (point_grid$start_peak_diff[i] == 0) {
      distance_s_3d[i] <- distance_s[i]
      incline_s[i] <- 0
    } else {
      # Calculate 3D distances
      distance_s_3d[i] <- sqrt(distance_s[i]^2 + point_grid$start_peak_diff[i]^2)
      incline_s[i] <- distance_s[i] / point_grid$start_peak_diff[i]
    }
    
    # Check for zero distances and handle them by assigning zero to distance_e_3d and incline_e
    if (distance_e[i] == 0) {
      distance_e_3d[i] <- 0
      incline_e[i] <- 0
    } else if (point_grid$end_peak_diff[i] == 0) {
      distance_e_3d[i] <- distance_e[i]
      incline_e[i] <- 0
    } else {
      # Calculate 3D distances
      distance_e_3d[i] <- sqrt(distance_e[i]^2 + point_grid$end_peak_diff[i]^2)
      incline_e[i] <- (distance_e[i] / point_grid$end_peak_diff[i]) * -1
    }
  }

# Add the distances to the dataframe
  point_grid$dist_total     <- distance_t
  point_grid$dist_start     <- distance_s
  point_grid$dist_end       <- distance_e
  point_grid$dist_start_3d  <- distance_s_3d
  point_grid$dist_end_3d    <- distance_e_3d
  point_grid$incline_start  <- incline_s
  point_grid$incline_end    <- incline_e
  
  
### UNTESTED VECTOR VERSION OF ABOVE:
## Calculate distances for each row
#point_grid$dist_total <- st_distance(point_start, point_end)
#point_grid$dist_start <- st_distance(point_start, point_peak)
#point_grid$dist_end <- st_distance(point_peak, point_end)
#
## Calculate 3D distances for each row, handling zero distances
#point_grid$dist_start_3d <- ifelse(point_grid$dist_start == 0, 0,
#                                   sqrt(point_grid$dist_start^2 + point_grid$start_peak_diff^2))
#point_grid$dist_end_3d <- ifelse(point_grid$dist_end == 0, 0,
#                                 sqrt(point_grid$dist_end^2 + point_grid$end_peak_diff^2))
#
## Calculate inclines for each row, handling zero distances
#point_grid$incline_start <- ifelse(point_grid$dist_start == 0, 0,
#                                    point_grid$dist_start / point_grid$start_peak_diff)
#point_grid$incline_end <- ifelse(point_grid$dist_end == 0, 0,
#                                  point_grid$dist_end / point_grid$end_peak_diff * -1)


# Calculate times for different helicopter movements on flat distances
  time_full_ascend_flat = (point_grid$start_peak_diff / 1000) / full_ascend_kmh
  time_full_level_flat = (point_grid$dist_total / 1000) / full_level_kmh
  time_full_descend_flat = (point_grid$end_peak_diff / 1000) / full_level_kmh
  time_empty_ascend_flat = (point_grid$end_peak_diff / 1000) / all_empty_kmh
  time_empty_level_flat = (point_grid$dist_total / 1000) / all_empty_kmh
  time_empty_descend_flat = (point_grid$start_peak_diff / 1000) / all_empty_kmh
  total_time_flat = round(
    (time_full_ascend_flat + time_full_level_flat + time_full_descend_flat +
     time_empty_ascend_flat + time_empty_level_flat + time_empty_descend_flat) * 60, 1
  ) + 2  # Adding 2 minutes for fill time

# Calculate times for different helicopter movements assuming flying on angles
  time_full_ascend_slope   = (point_grid$dist_start_3d / 1000) / full_ascend_kmh
  time_full_descend_slope  = (point_grid$dist_end_3d / 1000) / full_level_kmh
  time_empty_ascend_slope  = (point_grid$dist_end_3d / 1000) / all_empty_kmh
  time_empty_descend_slope = (point_grid$dist_start_3d / 1000) / all_empty_kmh
  total_time_slope = round(
    (time_full_ascend_slope + time_full_descend_slope +
     time_empty_ascend_slope + time_empty_descend_slope) * 60, 1
  ) + 2  # Adding 2 minutes for fill time

# Update 'total_time' and 'total_time_slope' columns in 'point_grid'
  point_grid$total_time_flat <- total_time_flat
  point_grid$total_time_slope <- total_time_slope
  
# Extract x, y coordinate values
  point_coords <- st_coordinates(st_as_sf(point_grid))

# Create ppp objects from the coordinates
  min_x <- min(point_coords[, 1])
  max_x <- max(point_coords[, 1])
  min_y <- min(point_coords[, 2])
  max_y <- max(point_coords[, 2])
  resolution <- 200
  window_spec <- owin(c(min_x, max_x), c(min_y, max_y))
  point_pattern_simple <- ppp(
    point_coords[, 1], 
    point_coords[, 2], 
    marks = point_grid$total_time_flat, 
    window = window_spec
    )
  point_pattern_slope <- ppp(
    point_coords[, 1], 
    point_coords[, 2], 
    marks = point_grid$total_time_slope, 
    window = window_spec
    )

# Perform IDW interpolation and make rasters
  idw_result <- idw(point_pattern_simple, power = 5, eps = resolution)
  idw_raster_simple <- as(idw_result, "RasterLayer")
  crs(idw_raster_simple) <- 3111
  
  idw_result <- idw(point_pattern_slope, power = 5, eps = resolution)
  idw_raster_slope <- as(idw_result, "RasterLayer")
  crs(idw_raster_slope) <- 3111

# Save the data frame, rasters and shapefiles
  saveRDS(point_grid, "point_grid_data.rds")
  st_write(point_grid,
           file.path(output_path, "point_grid.shp"), 
           overwrite=TRUE,
           append=FALSE)
  #near_lines<-st_as_sf(near_lines)
  #st_write(near_lines, 
  #         file.path(output_path, "near_lines.shp"),  
  #         overwrite=TRUE)
  writeRaster(idw_raster_simple, 
              file.path(output_path, "travel_time_simple.tif"), 
              format="GTiff", 
              overwrite=TRUE)
  writeRaster(idw_raster_slope, 
              file.path(output_path, "travel_time_slope.tif"), 
              format="GTiff", 
              overwrite=TRUE)




```

### Helicopter travel time map (simple)

-   Interactive map of time to closest reservoir (in minutes). Click on a point to see the minutes
-   This map assumes helicopters ascend and descend vertically, and fly horizontally in-between

```{r Chunk 7: Map 1 (simple), message=FALSE, warning=FALSE, out.width='100%'}

# Create a custom color palette from green (0) to red (30)
custom_palette <- colorRampPalette(c("green", "yellow", "orange", "red"))(10)
custom_palette <- colorBin(custom_palette, c(0, 30), 10, pretty = FALSE)

# Create a Leaflet map
map_raster_simple <- leaflet() %>%
  setView(-37.72, 147.20, zoom = 6) %>%
  addProviderTiles("OpenStreetMap") %>%
  addRasterImage(
    idw_raster_simple,
    project = TRUE,
    group = "Minutes",
    layerId = "Minutes",
    colors = custom_palette,
    opacity = 0.5
  ) %>%
  addPolygons(
    data = as(st_transform(wp_combined_poly, 4326), "Spatial"),
    stroke = FALSE,
    smoothFactor = 0.2,
    fillColor = 'blue',
    fillOpacity = 0.8
  ) %>%
  addImageQuery(
    idw_raster_simple,
    project = TRUE,
    layerId = "Minutes",
    digits = 1,
    prefix = "Helicopter return"
  ) %>%
  addLegend(
    "bottomright",
    values = range(values(idw_raster_simple)),
    pal = custom_palette,
    title = "Time in minutes",
    opacity = 0.5
  )

map_raster_simple


```

### Helicopter travel time map (alternative)

-   This map assumes helicopters fly on angles over peaks

```{r Chunk 8: Map 2 (advanced), message=FALSE, warning=FALSE, out.width='100%'}

# Create a custom color palette from green (0) to red (30)
custom_palette <- colorRampPalette(c("green", "yellow", "orange", "red"))(10)
custom_palette <- colorBin(custom_palette, c(0, 30), 10, pretty = FALSE)

# Create a Leaflet map
map_raster_slope <- leaflet() %>%
  setView(-37.72, 147.20, zoom = 6) %>%
  addProviderTiles("OpenStreetMap") %>%
  addRasterImage(
    idw_raster_slope,
    project = TRUE,
    group = "Minutes",
    layerId = "Minutes",
    colors = custom_palette,
    opacity = 0.5
  ) %>%
  addPolygons(
    data = as(st_transform(wp_combined_poly, 4326), "Spatial"),
    stroke = FALSE,
    smoothFactor = 0.2,
    fillColor = 'blue',
    fillOpacity = 0.8
  ) %>%
  addImageQuery(
    idw_raster_slope,
    project = TRUE,
    layerId = "Minutes",
    digits = 1,
    prefix = "Helicopter return"
  ) %>%
  addLegend(
    "bottomright",
    values = range(values(idw_raster_slope)),
    pal = custom_palette,
    title = "Time in minutes",
    opacity = 0.5
  )

map_raster_slope


```

### More sophisticated helicopter travel times

-   Build lines between water points and demand points
-   Point sample every 100m
-   Attribute each point with linear distance from water point (x) and elevation (z)
-   Draw a convex hull around points
-   Convert to points (if not already)
-   Delete any points falling where z \< a line between wp and dp (in case of -v- shape)
-   Check ordering of points is still intact
-   Calculate travel times between point pairs (0, 1 to n-1, n)
-   Start helicopter speed at 0 km/h at water point with easing
-   Slow helicopter to (?) km/h at demand point
-   all other travel speeds proportional to slope & load
-   Sum each path
-   Create rasters using IDW (distance-weighted smoothing) for efficient maps

\`\`\`{r Chunk 9: Better helicopter travel times, message=FALSE, warning=FALSE} knitr::opts_chunk\$set(cache = TRUE)
